---
title: "Homework 1"
author: "Marco Camalich"
date: 'Assigned: January 18, 2017'
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: cerulean
    highlight: tango
---

##### Citation

This homework assignment is largely borrowed from Coursework created Prof. Alexandra Chouldechova at Carnegie Mellon University.  I have modified slightly and am using in accordance with the Creative Commons License:

 
Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License(http://creativecommons.org/licenses/by-nc-sa/4.0/)

##### This homework is due by **11:55PM on Friday, Sep 22**.  

##### To complete this assignment, follow these steps:

1. Download the `homework1.Rmd` file from Moodle

2. Open `homework1.Rmd` in RStudio.

3. Replace the "Your Name Here" text in the `author:` field with your own name.

4. Supply your solutions to the homework by editing `homework1.Rmd`.

5. When you have completed the homework and have **checked** that your code both runs in the Console and knits correctly when you click `Knit HTML`, rename the R Markdown file to `homework1_YourNameHere.Rmd`, and submit both the `.Rmd` file and the `.html` output file on Blackboard.  (YourNameHere should be changed to your own name.)

##### Homework tips:

1. Recall the following useful RStudio hotkeys.

Keystroke | Description
------------|-------------------------------------------
`<tab>` | Autocompletes commands and filenames, and lists arguments for functions.
`<up>` | Cycles through previous commands in the console prompt
`<ctrl-up>` | Lists history of previous commands matching an unfinished one
`<ctrl-enter>` | Runs current line from source window to Console. Good for trying things out ideas from a source file.
`<ESC>` | Aborts an unfinished command and get out of the + prompt

**Note**: Shown above are the Windows/Linux keys.  For Mac OS X, the `<ctrl>` key should be substituted with the `<command>` (&#8984;) key.

2. Instead of sending code line-by-line with `<ctrl-enter>`, you can send entire code chunks, and even run all of the code chunks in your .Rmd file. Look under the <Chunks> menu of the Source panel.

3. Run your code in the Console and Knit HTML frequently to check for errors.

4. You may find it easier to solve a problem by interacting only with the Console at first, or by creating a separate `.R` source file that contains only R code and no Markdown.

### Housekeeping - packages to be used.

We will give preference to the "tidyverse" of packages related to "Tidy Data."  


```{r}
library(knitr)
library(tidyverse) # loads ggplot2, dplyr,tidyr,readr,purr,tibble
library(broom)  # because I find it useful
library(lattice)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")  

options(scipen = 4)  # make output prefer not to use scientific notation
```
### Introduction: Bikeshare data

For this problem we'll be working with two years of bikeshare data from the Capital Bikeshare system in Washington DC.  The dataset contains daily bikeshare counts, along with daily measurements on environmental and seasonal information that may affect the bikesharing.  

### Data pre-processing 

Let's start by loading the data.

```{r}
bikes <- read.csv("http://www.andrew.cmu.edu/user/achoulde/95791/data/bikes.csv", header = TRUE)

# Transform temp and atemp to degrees C instead of [0,1] scale
# Transform humidity to %
# Transform wind speed (multiply by 67, the normalizing value)

bikes <- mutate(bikes,
                   temp = 47 * temp - 8,
                   atemp = 66 * atemp - 16,
                   hum = 100 * hum,
                   windspeed = 67 * windspeed)

# The recode() command from the dplyr library allows us to easily
# rename values in our variables.  Below we use this command to change season
# from numeric codings to season names.

bikes <- mutate(bikes, 
                   season = recode(season, `1`="Winter",`2`="Spring",
                                      `3`="Summer", `4`="Fall"))
bikes
```

Let's look at some boxplots of how bikeshare ride count varies with season.

```{r, fig.height = 4, fig.width = 5} 
qplot(data = bikes, x = season, y = cnt, fill = I(cbPalette[3]), geom = "boxplot")
```

There's something funny going on here.  Instead of showing up in seasonal order, the seasons in the plot are showing up in **alphabetical order**.  The following command reorders the seasons appropriately.

```{r}
bikes <- mutate(bikes, season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")))
```

Now let's try that plot again.

```{r, fig.height = 4, fig.width = 5} 
qplot(data = bikes, x = season, y = cnt, fill = I(cbPalette[3]), geom = "boxplot")
```

Here's information on what the variables mean.

  - instant: record index
	- dteday : date
	- season : season (1:Winter, 2:Spring, 3:Summer, 4:Fall)
	- yr : year (0: 2011, 1:2012)
	- mnth : month ( 1 to 12)
	- hr : hour (0 to 23)
	- holiday : weather day is holiday or not (extracted from http://dchr.dc.gov/page/holiday-schedule)
	- weekday : day of the week
	- workingday : if day is neither weekend nor holiday is 1, otherwise is 0.
	+ weathersit : 
		- 1: Clear, Few clouds, Partly cloudy, Partly cloudy
		- 2: Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist
		- 3: Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds
		- 4: Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog
	- temp : Temperature in Celsius. 
	- atemp: `feels like' temperature in Celsius. 
	- hum: Normalized humidity. The values are divided to 100 (max)
	- windspeed: Normalized wind speed. The values are divided to 67 (max)
	- casual: count of casual users
	- registered: count of registered users
	- cnt: count of total rental bikes including both casual and registered

### Problem 1: Qualitative predictors

> The Season variable is an example of what's called a *qualitative* or *categorical* predictor.  In R, such variables are called `factors`.  This problems gets to fit a model with a qualitative predictor and to interpret the findings.


##### **(a)** Suppose we were interested in predicting the count of total bike rentals by using the *season* as the predictor.  Fit a linear (using the lm command) regression model with `cnt` as the response and `season` as the input.  Use the `tidy()` and `kable()` commands to produce a nice looking coefficients table.

(Maybe look at https://rstudio-pubs-static.s3.amazonaws.com/58870_136d938b1ffd427ab5b30a98b663b127.html  for an idea)
```{r}
unique(bikes[3])
```

```{r}
model <- lm(cnt ~ season, data = bikes)
coef <- tidy(model)
kable(coef, caption = "Coefficients of Linear Regression Model")

```

##### **(b)** How many total coefficients are there in the model?

- **5. The 4 seasons and the intercept.**
    
    
##### **(c)** How many coefficients are estimated for the `season` variable?
  
- **3, as there are 4 levels and Winter is used as the reference level for the other 3.**
    
    
##### **(d)** Interpret the coefficients of `season` in the model.
   
- **Intercept (2604.133): This is the estimated count of bike rentals during Winter when all other factors are the same. It's like the baseline number of rentals in the coldest season.** 

- **SeasonSpring (2388.199): Compared to Winter, there are approximately 2388 more bike rentals in Spring when everything else is constant. We can conclude that people tend to rent more bikes in Spring than in Winter.**

- **SeasonSummer (3040.171): In Summer, the bike rentals increase by around 3040 compared to Winter, assuming all other factors remain constant. We can conclude that summer sees the highest bike rentals among all seasons.**

- **SeasonFall (2124.030): Fall shows an increase of about 2124 rentals compared to Winter, assuming everything else stays the same. We can conclude that fall is also a popular time for bike rentals, although not as high as Summer.**

<p> **Hint**: If you have not previously studied how to interpret qualitative variables in regressions, perhaps search on key terms 'lm', 'r', 'categorical predictor' 'interpreting'. </p>

<hr>

### Problem 2: Multiple linear regression

> In this problem we'll practice fitting and interpreting the results of a multiple linear regression.

##### **(a)** Fit a regression model with `cnt` as the response and the following variables as inputs: `temp`, `atemp`, `mnth`, `hum`, `windspeed`.  Produce a nice looking coefficients table.

```{r}
model2 <- lm(cnt ~ temp + atemp + mnth + hum + windspeed, data = bikes)
coef2 <- tidy(model2)
kable(coef2, caption = "Coefficients of Multiple Linear Regression Model")
```

##### **(b)** Interpret the coefficients of `temp`, `mnth` and `hum` in the model.

- **The coefficient for temp is approximately 45.4. It means that for every 1-degree Celsius increase in temperature, there's an average increase of 45.4 bike rentals.**
- **The coefficient for mnth is approximately 95.0. It means that there is an increase of around 95 bike rentals for every unit increase in the month.**
- **The coefficient for hum is approximately -35.3. It means that for every 1% increase in humidity, there is a decrease of 35.3 bike rentals.**
    
    
##### **(c)** Which predictors are associated with increased ridership?  Which predictors are associated with decreased ridership?
  
- **Humidity and wind speed are associated with decreased ridership while temperature, "feels like" temperature and the month of the year are associated with increased ridership.**
    
##### **(d)** Which predictors are statistically significant at the 0.05 level?
   
- **Month, humidity and windspeed.**

<hr>

### Problem 3:  Dealing with collinearity 

> As you probably already know from your most recent regression class, *collinear* or *highly correlated* predictors can make interpreting regression coefficients problematic.  In this problem you will try to diagnose and address collinearity issues in the data.

##### **(a)** Use the `pairs()` function on the set of variables used in **Problem 2** to check if any of the predictor variables are highly correlated with one another.  Your pairs plot should have scatterplots above the diagonal, and correlations below the diagonal.

```{r}
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y))
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste0(prefix, txt)
  if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor, col = "black")
}

pairs(bikes[c("temp", "atemp", "mnth", "hum", "windspeed")], lower.panel = panel.cor)
```

**Hint**: A complete example of how to use the `pairs()` command to construct such plots may be found here: [Pairs plot example](http://www.andrew.cmu.edu/user/achoulde/94842/lectures/lecture08/lecture08-94842.html#collinearity-and-pairs-plots)

##### **(b)** Are any of the predictors highly correlated?  Are you surprised that these predictors are highly correlated, or can you think of a reason for why it makes sense that they should be correlated?

- **The only highly correlated predictors are temp and atemp as they are calculatin essentially the same. I'm not surprised as the humidity, wind speed and month can vary, but the temperatures feeling and the actual temperature will feel similar, if not equally the same.**

##### **(c)** Refit your regression model, but this time **omit** the `temp` variable.  Display the coefficients table for this model.

```{r}
model3 <- lm(cnt ~ atemp + mnth + hum + windspeed, data = bikes)
coef3 <- tidy(model3)
kable(coef3, caption = "Coefficients of Multiple Linear Regression Model")
```

##### **(d)** What is the coefficient of `atemp` in this new model?  Is it very different from the `atemp` coefficient estimated in part **(b)**?  Is it statistically significant?  Explain your findings.

- **The new coefficient of atemp is 108.2. It is significantly different than the first one as the other was 72.01. It became very statistically significant as its p-value lowered to 0, making it practically exact.**

# Optional Material Below

### Problem 4: Exploring non-linearities

> **Hint**: For this problem, you will find it useful to know about the `jitter` feature in graphics.  [Begin by reviewing the code at this link](http://www.andrew.cmu.edu/user/achoulde/94842/misc/extra_tips.html#jittering-points), and be sure to use what you feel to be an appropriate amount of jitter in your plots for **(a)**, **(b)** and **(c)**.  You **should not** use jitter for parts **(d)** onward.  

##### **(a)** Using `ggplot2` graphics, construct a scatterplot of `cnt` (bikeshare count) across `mnth` (month of the year).  Describe what you see.  Does a linear relationship appear to be a good way of modeling how bikeshare count varies with month?  

```{r}
ggplot(bikes, aes(x = mnth, y = cnt)) +
  geom_jitter(width = 0.3, height = 0, color = "#0072B2", alpha = 0.7) +
  labs(x = "Month", y = "Bikeshare Count") +
  theme_minimal()

```

- **Looking at the plot, we can conclude that a linear relationship might not be the best way to model how bikeshare counts vary with the month of the year. The data points do not follow a clear linear pattern. We can see there are fluctuations and variations in bikeshare counts across different months**

##### **(b)** Use `ggplot2`'s `stat_smooth()` overlays to try out *different degree polynomial fits* for modeling the relationship between `cnt` and `month`.  Display the lowest degree polynomial fit that appears to nicely capture the trends in the data.  Explain your choice.

```{r}
ggplot(bikes, aes(x = mnth, y = cnt)) +
  geom_jitter(width = 0.3, height = 0, color = "#0072B2", alpha = 0.7) +
  stat_smooth(method = "lm", formula = y ~ poly(x, degree = 1), se = FALSE, color = "#D55E00", linewidth = 1) +
  stat_smooth(method = "lm", formula = y ~ poly(x, degree = 2), se = FALSE, color = "#56B4E9", linewidth = 1) +
  stat_smooth(method = "lm", formula = y ~ poly(x, degree = 3), se = FALSE, color = "#009E73", linewidth = 1) +
  labs(x = "Month", y = "Bikeshare Count") +
  theme_minimal() +
  scale_color_manual(values = c("#D55E00", "#56B4E9", "#009E73")) +
  guides(color = guide_legend(title = "Polynomial Degree"))

```

- **The best line is the green one as it is not as high as the next degrees but still works as a predictor.**

